---
title: "RNA & ADT Expression of Cells with Overlapping TCRs"
author: "John Moore"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(dplyr)
library(ggplot2)
library(Seurat)
library(here)
library(patchwork)
library(gridExtra)
library(pheatmap)
library(DT)
library(RColorBrewer)
here::i_am("stim_TCRs/TCR_summaries/scripts/RNA_expression_of_overlapping_TCRs.R")

path= "/stor/work/Ehrlich_COVID19/analysis/seurat/Qi_clustering/stimulatedTCR/stimuT_clusters_harmony.rds"
stimTCR_obj= readRDS(path)
plot_vec= c(11,7)
```


```{r}
##-----------
# Adding TRUST4 TCR meta.data
##-----------
meta_data= stimTCR_obj@meta.data
path= here("stim_TCRs","data","overlapping_TCRs","stimTCR_for_metadata.csv")
TRUST4_TCR_data= read.csv(path)
meta_data= left_join(meta_data, TRUST4_TCR_data, by= "barcode")
stimTCR_obj@meta.data= meta_data
```


```{r}
##-----------
# Subsetting overlapping TCRs
##-----------
path= here("stim_TCRs","data","overlapping_TCRs","all_overlapping_stimTCRs.csv")
all_overlapping_stimTCRs= read.csv(path) %>% pull(x)
path= here("stim_TCRs","data","overlapping_TCRs","overlap_unstim_TCRs.csv")
overlap_unstim_TCRs= read.csv(path) %>% pull(x)
stimTCR_obj@meta.data$is_overlap= stimTCR_obj@meta.data$TCR %in% all_overlapping_stimTCRs
stimTCR_obj@meta.data$is_unstim_overlap= stimTCR_obj@meta.data$TCR %in% overlap_unstim_TCRs
  ## Adding overlap data to visualize on UMAP & subset
stimTCR_MD= stimTCR_obj@meta.data %>% filter(!is.na(TCR))
# path= here("stim_TCRs","data","stimTCRs_noOverlap.csv")
# write.csv(stimTCR_MD, path, row.names= F)

overlap_obj= stimTCR_obj ## copying stimTCR_obj
overlap_obj@meta.data = stimTCR_obj@meta.data[stimTCR_obj$is_overlap,]
  ## Manually overwriting the meta.data, thus subsetting the obj b/c subset() wasn't working.
  ## Keeping only overlapping TCRs

unstim_overlap_obj= stimTCR_obj
unstim_overlap_obj@meta.data= stimTCR_obj@meta.data[stimTCR_obj$is_unstim_overlap,] 
  ## Keeping only unstimulated overlapping TCRs
```



```{r, fig.dim= plot_vec}
##-----------
# Visualizing Overlapping TCRs on UMAP ADT & RNA Expression Profile
##-----------
p1= DimPlot(stimTCR_obj,cells.highlight= stimTCR_obj$barcode[stimTCR_obj$is_overlap]) + 
  NoLegend() + 
  ggtitle("All \n Overlapping TCRs")
p2= DimPlot(stimTCR_obj,cells.highlight= stimTCR_obj$barcode[stimTCR_obj$is_unstim_overlap]) + 
  NoLegend() + 
  ggtitle("Unstimulated \n Overlapping TCRs")
p3= DimPlot(stimTCR_obj,cells.highlight= stimTCR_obj$barcode[(stimTCR_obj$is_overlap & stimTCR_obj$is_unstim_overlap == F)]) + 
  NoLegend() + 
  ggtitle("Other \n Overlapping TCRs")

p1 | p2 | p3
```


```{r,plot_vec, message= F, warning= F, fig.dim= plot_vec}
DimPlot(stimTCR_obj, split.by= "severity", group.by= "severity")
DimPlot(stimTCR_obj, split.by= "calc_age_bin", group.by= "calc_age_bin")
DimPlot(stimTCR_obj, group.by= "antigen", split.by= "antigen")

## Regression Model of the overlap by age 
regression_data= stimTCR_MD %>% 
  group_by(subject_id) %>%
  mutate(prop_overlap= round(sum(is_overlap)/n(),digits= 2),
         prop_unstim_overlap= round(sum(is_unstim_overlap)/n(),digits= 2)) %>%
  ungroup()

overlap_model= lm(data= regression_data, formula= prop_overlap ~ age)
summary(overlap_model)

r1= ggplot(regression_data, aes(x= age, y= prop_overlap, color= severity)) + 
  geom_point() +
  geom_smooth(method= "lm", formula= y ~ x, color= "black") +
  annotate("text",
           label= paste0(round(overlap_model$coefficients[2], digits= 4),
                         "*age (years)"),
           x= 30, y= 0.4) +
  ggtitle("Adjusted R^2 value of 0.018")



unstim_overlap_model= lm(data= regression_data, 
                         formula= prop_unstim_overlap ~ age)
summary(unstim_overlap_model)
r2= ggplot(regression_data, aes(x= age, y= prop_unstim_overlap, color= severity)) +
  geom_point() + 
  geom_smooth(method= "lm", formula= y ~ x, color= "black") + 
  annotate("text",
           label= paste0(round(unstim_overlap_model$coefficients[2], digits= 4),
                         "*age (years)"),
           x= 30, y= 0.4) +
  ggtitle("Adjusted R^2 value of 0.064")
r1 | r2

regression_data %>%
  distinct(subject_id, severity, calc_age_bin, prop_overlap, prop_unstim_overlap) %>% 
  datatable

overlap_model_w_age= lm(data= regression_data, 
                        formula= prop_overlap ~ age * severity)
summary(overlap_model_w_age)

unstim_overlap_model_w_age= lm(data= regression_data, 
                        formula= prop_unstim_overlap ~ age * severity)
summary(unstim_overlap_model_w_age)


sev_plot_data= stimTCR_MD %>% 
  group_by(severity) %>% 
  mutate(overlap_count= sum(is_overlap),
         prop_overlap_sev= overlap_count/n()) %>%
  distinct(severity, overlap_count, prop_overlap_sev)
ggplot(sev_plot_data, aes(x= severity, y= prop_overlap_sev, fill= severity)) +
  geom_col()
ggplot(sev_plot_data, aes(x= severity, y= overlap_count, fill= severity)) + 
  geom_col()

sev_age_plot_data= stimTCR_MD %>% 
  group_by(severity, calc_age_bin) %>% 
  mutate(overlap_count= sum(is_overlap),
         prop_overlap= overlap_count/n(),
         calc_age_bin= factor(calc_age_bin, levels= c("Pediatric","Adult","Geriatric"))) %>% 
  distinct(severity, calc_age_bin, overlap_count, prop_overlap)
ggplot(sev_age_plot_data, aes(x= severity, y= prop_overlap, fill= calc_age_bin)) +
  geom_col(position= "dodge") 
ggplot(sev_age_plot_data, aes(x= severity, y= overlap_count, fill= calc_age_bin)) +
  geom_col(position= "dodge")
```


<!-- It looks like age is significant, but it has a super small effect, and the slope seems to be driven in part by the big outliers.

The larger number and percentage of outliers in severe individuals is consistent with the idea that direct ex vivo stimulated T cells are the source of the overlapping TCRs. (Direct ex vivo stimulated meaning that the T cells were activated in vivo.) -->

## Overlapping RNA Expression Profile

- Genes ordered by AUC  

```{r, fig.dim= c(11,30), message= F, warning= F}
##-----------
# Overlapping RNA Expression Profile
##-----------
DefaultAssay(stimTCR_obj) <- "RNA"

Idents(stimTCR_obj)= "is_overlap"
# overlap_genes_df= FindAllMarkers(stimTCR_obj, test.use= "roc") %>%
#   filter(cluster == F) %>%
#   arrange(desc(myAUC))
  ## use roc as opposed to wilcox
  ## Clusters are doubled b/c compared against a control group
path= here("stim_TCRs","data","overlap_genes_df.csv")
# write.csv(overlap_genes_df,
#           path,
#           row.names = F)
overlap_genes_df= read.csv(path)
position_data= as.data.frame(stimTCR_obj[["wnn.umap"]]@cell.embeddings)
position_data$barcode= row.names(position_data)
stimTCR_MD= left_join(stimTCR_MD, position_data, by= "barcode")
# path= here("stim_TCRs","data","stimTCR_w_overlapping.csv")
# write.csv(stimTCR_MD, path, row.names= F)


overlap_features= overlap_genes_df$gene
overlap_features= c("LTB", "TNFRSF4", "TXN", "CD40LG", "TMSB10", "PFN1", 
                    "PRDX1", "TNFRSF18", "GZMB", "PTPRC", "GZMH", "PRF1",
                    "GNLY","CST7", "CCL5")
  
stimFeat.list <- list()

overlap_MD= stimTCR_MD %>% filter(is_overlap)
RNA_data= stimTCR_obj@assays[["RNA"]]@data %>%
  as.data.frame() %>% 
  t() %>% 
  as.data.frame() %>%
  filter(row.names(.) %in% overlap_MD$barcode)

for(feat in overlap_features){
  overlap_MD$tmp= RNA_data[,feat]
  stimFeat.list[[feat]]= ggplot(overlap_MD) +
      aes(x= wnnUMAP_1, y= wnnUMAP_2, color= tmp) +
      geom_point(size= 1) +
      scale_color_viridis_c() + 
      ggtitle(feat)
}
wrap_plots(stimFeat.list, ncol= 3)
```

```{r, fig.dim= plot_vec, message= F, warning= F}

  ## I'm pretty sure this is showing all the cells that express that gene.
  ## I could make this more specific w/ Dimplots and highligt cells.
  ## Not sure if I want to. 

## Lympotoxin beta, TNFRSF4, TXN, CD40LG, TMSB10, PFN1, 

## LTB     - lymphotoxin beta
## TNFRSF4 - aka CD134/OX40 (This is super interesting.)
## TXN     - thioredoxin
## CD40LG  - different than CD40L? expressed on activated T cellls 
## TMSB10  - beta-thymosin, not sure the relevance beyond actin remodeling and some influence w/ M1 macrophages
## PFN1    - regulates actin polymerization (Related to TMSB10?)
## MYL6    - myosin light polypeptide 6 
## RPS     - ribosomal protein group S
## RPL     - ribosomal protein group L 


## PRDX1    - Peroxiredoxin 1 (may be proliferative or antiviral for CD8 T cells)
## ACTB     - beta-actin 
## PSME2    - Proteasome activator complex subunit 2 
## TNFRSF18 - (Big AUC drop after this gene)
## GZMB     - Granzyme B
## HLA-C    - 
## PTPRC    - aka CD45 
## CD52     - 
## BTG1     - active in T cell quiescence 

## CD8A    - 
## MALAT1  -
## GZMH    - 
## PRF1    -
## ZFP36L2 - 
## HLA-A   - 
## GNLY    - 
## CST7    -
## NKG7    - 
## MT-CO1  -
## CCL5    -


## Which group is upregulated where? 

heatmap_data= overlap_obj@assays[["RNA"]]@data %>% 
  as.data.frame() %>% 
  filter(row.names(.) %in% overlap_genes_df$gene)
heatmap_data= heatmap_data[,colnames(heatmap_data) %in% overlap_obj@meta.data$barcode] %>% 
              as.matrix
hm_palette <- rev(colorRampPalette(colors = c("yellow","black","magenta"))(100))

pheatmap(heatmap_data,
         cluster_cols= F,
         cluster_rows= F,
         show_colnames= F,
         color= hm_palette,
         main= "RNA expression of cells w/overlapping TCRs")
  ## The issue with DoHeatmap is that the subsetting wasn't working b/c 
  ## the dim of meta.data and RNA assay was not the same. 
```


## Overlapping ADT Expression Profile
```{r, fig.dim= plot_vec, message= F, warning= F}
##-----------
# Overlapping ADT Expression Profile
##-----------
DefaultAssay(stimTCR_obj) <- "ADT"
Idents(stimTCR_obj)= "is_overlap"

# overlap_ADT_df= FindAllMarkers(stimTCR_obj, test.use= "roc") %>%
#   filter(cluster == F) %>% 
#   arrange(desc(myAUC))
path= here("stim_TCRs","data","overlap_ADT_df.csv")
#write.csv(overlap_ADT_df, path, row.names = F)
overlap_ADT_df= read.csv(path)

DefaultAssay(overlap_obj)= "ADT"
ADT_data= stimTCR_obj@assays[["ADT"]]@counts %>%
  as.data.frame() %>% 
  t() %>% 
  as.data.frame() %>%
  filter(row.names(.) %in% overlap_MD$barcode)

overlap_ADTs= overlap_ADT_df$gene
stimADT.list <- list()


for(feat in overlap_ADTs){
  overlap_MD$tmp= ADT_data[,feat]
  stimADT.list[[feat]]= ggplot(overlap_MD) +
      aes(x= wnnUMAP_1, y= wnnUMAP_2, color= tmp) +
      geom_point(size= 1) +
      scale_color_viridis_c() + 
      ggtitle(feat)
}
wrap_plots(stimADT.list, ncol= 3)
```


```{r, fig.dim= plot_vec, message= F, warning= F}
heatmap_data= overlap_obj@assays[["ADT"]]@data %>% 
  as.data.frame() %>% 
  filter(row.names(.) %in% overlap_ADT_df$gene)
heatmap_data= heatmap_data[,colnames(heatmap_data) %in% overlap_obj@meta.data$barcode] %>% 
  as.matrix
hm_palette <- rev(colorRampPalette(colors = c("yellow","black","magenta"))(100))

pheatmap(heatmap_data,
         cluster_cols= F,
         cluster_rows= F,
         show_colnames= F,
         color= hm_palette,
         main= "ADT expression of cells w/overlapping TCRs")
```


## Cells with TCRs that overlap w/ Unstimultaed RNA Expression Profile
```{r, fig.dim= c(11,30), warning= F, message= F}
##-----------
# TCRs Overlapping w/ Unstimultaed RNA Expression Profile
##-----------
DefaultAssay(stimTCR_obj)= "RNA"
Idents(stimTCR_obj) = "is_unstim_overlap" 
# unstim_overlap_genes_df= FindAllMarkers(stimTCR_obj, test.use= "roc") %>%
#   filter(cluster == F) %>% 
#   arrange(desc(myAUC))
path= here("stim_TCRs","data","unstim_overlap_genes_df.csv")
#write.csv(unstim_overlap_genes_df, path, row.names= F) 
unstim_overlap_genes_df= read.csv(path)

unstim_overlap_RNAs= unstim_overlap_genes_df$gene
unstim_overlap_RNAs= c("LTB","TNFRSF4","CD40LG","TXN","MAL","TMSB10","PFN1",
                       "PRDX1","PTPRC","HOPX","CCL4","GZMB","GZMH","PRF1","GNLY",
                       "CCL5","NKG7")
  ## subsetting b/c too many 
  ## A lot of ribosomal proteins were significant for some reason
  ## Would that be an indicator of protein translation due to activation? 
unstim_overlapRNA.list <- list()
unstim_overlap_MD= stimTCR_MD %>% filter(is_overlap)

unstim_RNA_data= stimTCR_obj@assays[["RNA"]]@counts %>%
  as.data.frame() %>% 
  t() %>% 
  as.data.frame() %>%
  filter(row.names(.) %in% unstim_overlap_MD$barcode)

for(feat in unstim_overlap_RNAs){
  unstim_overlap_MD$tmp= unstim_RNA_data[,feat]
  unstim_overlapRNA.list[[feat]]= ggplot(unstim_overlap_MD) +
      aes(x= wnnUMAP_1, y= wnnUMAP_2, color= tmp) +
      geom_point(size= 1) +
      scale_color_viridis_c() + 
      ggtitle(feat)
}
wrap_plots(unstim_overlapRNA.list, ncol = 3)
```

A lot of ribosomal proteins were significant for some reason. Would that be an indicator of protein translation due to activation? 

```{r, fig.dim= plot_vec, message= F, warning= F}
heatmap_data= unstim_overlap_obj@assays[["RNA"]]@data %>% 
  as.data.frame() %>%
  filter(row.names(.) %in% unstim_overlap_genes_df$gene)
heatmap_data= heatmap_data[,colnames(heatmap_data) %in% overlap_obj@meta.data$barcode] %>% 
  as.matrix
hm_palette <- rev(colorRampPalette(colors = c("yellow","black","magenta"))(100))

pheatmap(heatmap_data[1:ceiling(nrow(heatmap_data)/2),],
         cluster_cols= F,
         cluster_rows= F,
         show_colnames= F,
         color= hm_palette,
         main= "RNA expression of cells w/ TCRs that overlap w/ unstimulated TCRs")
  ## First half of the genes

pheatmap(heatmap_data[ceiling(nrow(heatmap_data)/2):nrow(heatmap_data),],
         cluster_cols= F,
         cluster_rows= F,
         show_colnames= F,
         color= hm_palette,
         main= "RNA expression of cells w/ TCRs that overlap w/ unstimulated TCRs")
  ## 2nd half of the genes
```

## Cells with TCRs that overlap w/ Unstimultaed ADT Expression Profile
```{r, fig.dim= plot_vec, warning= F, message= F}
##-----------
# TCRs Overlapping w/ Unstimultaed ADT Expression Profile
##-----------
DefaultAssay(stimTCR_obj) <- "ADT"
Idents(stimTCR_obj)= "is_unstim_overlap"

# unstim_overlap_ADT_df= FindAllMarkers(stimTCR_obj, test.use= "roc") %>%
#   filter(cluster == F) %>% 
#   arrange(desc(myAUC))
path= here("stim_TCRs","data","unstim_overlap_ADT_df.csv")
# write.csv(unstim_overlap_ADT_df, path, row.names= F)
unstim_overlap_ADT_df= read.csv(path)

DefaultAssay(unstim_overlap_obj)= "ADT"
unstim_ADT_data= stimTCR_obj@assays[["ADT"]]@counts %>%
  as.data.frame() %>% 
  t() %>% 
  as.data.frame() %>%
  filter(row.names(.) %in% unstim_overlap_MD$barcode)

unstim_overlap_ADTs= unstim_overlap_ADT_df$gene
overlap_unstimADT.list <- list()


for(feat in unstim_overlap_ADTs){
  overlap_MD$tmp= ADT_data[,feat]
  overlap_unstimADT.list[[feat]]= ggplot(overlap_MD) +
      aes(x= wnnUMAP_1, y= wnnUMAP_2, color= tmp) +
      geom_point(size= 1) +
      scale_color_viridis_c() + 
      ggtitle(feat)
}
wrap_plots(overlap_unstimADT.list, ncol= 3)






heatmap_data= unstim_overlap_obj@assays[["ADT"]]@data %>% 
  as.data.frame() %>%
  filter(row.names(.) %in% unstim_overlap_ADT_df$gene)
heatmap_data= heatmap_data[,colnames(heatmap_data) %in% overlap_obj@meta.data$barcode] %>% 
  as.matrix
hm_palette <- rev(colorRampPalette(colors = c("yellow","black","magenta"))(100))

pheatmap(heatmap_data,
         cluster_cols= F,
         cluster_rows= F,
         show_colnames= F,
         color= hm_palette,
         main= "ADT expression of cells w/ TCRs that overlap w/ unstimulated TCRs")
```

### Overlapping TCRs by Antigen Group
```{r, warning= F, message= F}
##-----------
# Overlapping TCRs by Antigen Group
##-----------
plot.list <- list()
for(antigen in unique(stimTCR_obj$antigen)){
  plot.list[[antigen]]= DimPlot(stimTCR_obj, 
                                cells.highlight= stimTCR_obj$barcode[stimTCR_obj$antigen == antigen &
                                                                     stimTCR_obj$is_overlap]) + 
                        NoLegend() + 
                        ggtitle(paste0(antigen))
}
wrap_plots(plot.list, ncol= 3)


print("REMOVED TCRs") 
removed= stimTCR_obj@meta.data %>% 
  filter(is_overlap == T) %>% 
  count(antigen) 
total_counts= stimTCR_obj@meta.data %>% count(antigen) %>% pull(n)
removed %>% mutate(`Proportion Removed`= round(removed$n/total_counts, digits= 2))

print("REMAINING TCRs")
remaining= stimTCR_obj@meta.data %>% 
  filter(is_overlap == F) %>% 
  count(antigen) 
remaining 
```



